---
title: "discussion13"
author: "Keith Colella"
date: "2023-11-25"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

# Question

A set of plastic spheres are to be made with a diameter of 1cm. If the manufacturing process is accurate to 1mm, what is the propagated error in volume of the spheres? Use differentials to approximate propagated error.

# Response

The volume of a sphere is defined as $$\frac{4}{3} \pi r^3$$. The accuracy of 1mm represents the differential for change in diameter, which we can denote as $$dd$$. Our goal is find the resulting change in volume, which we'll denote $$dv$$. We can do this by finding the derivative $$\frac{dv}{dd}$$.

First, we express $$v$$ in terms of $$d$$.

$$v = \frac{4}{3} \pi \frac{d}{2}^3$$

Then, we differentiate $$v$$ with respect to $$d$$. We set this up as $$\frac{dv}{dd}$$, and substitute $$v$$ with the volume formulation above.

$$\frac{dv}{dd} = \frac{d \frac{4}{3} \pi \frac{d}{2}^3}{dd}$$

We can write this more cleanly as follows.

$$\frac{dv}{dd} = \frac{d}{dd} (\frac{4}{3} \pi \frac{d}{2}^3)$$

We can treat $$\frac{4}{3} \pi$$ as a constant and really focus on differentiating $$\frac{d}{2}^3$$, which we can break into two functions: (1) $$f(x) = x^3$$ and (2) $$g(y) = \frac{y}{2}$$. If we apply the chain rule, we first find $$f'(x) = 3x^2$$, then find $$g'(y) = 1/2$$, and take the product of these two. This simplifies our derivative to the following.

$$\frac{dv}{dd} = \frac{4}{3} \pi \times 3\frac{d}{2}^2 \times \frac{1}{2}$$

Next, we evaluate at $$d = 1$$ (because our diameter is 1cm) and $$dd = 0.1$$ (because the change in diamater is 1mm, or 0.1cm).

$$\frac{dv}{dd} = \frac{4}{3} \pi \times 3\frac{1}{2}^2 \times \frac{1}{2}$$

This simplifies to ~1.571, which indicates that, for a 1cm change in diameter, our volume changes by ~1.571cm. However, we are looking for the change in volume resulting from a 1*mm* change in diamater. So, we convert this change to 0.1cm, then multiply it by ~1.571 to get ~0.1571.



```{r}
4 / 3 * pi * 3 * (1/2)^2 * (1/2)
```


```{r}
# Define the function for the derivative of the volume with respect to diameter
dv_dd <- function(d) {
  (4/3) * pi * 3 * (d/2)^2 * (1/2)
}

# Given values
d <- 1.0  # diameter in cm
dd <- 0.1  # change in diameter in cm (accuracy)

# Calculate the propagated error in volume
propagated_error <- dv_dd(d) * dd

# Print the propagated error
propagated_error
```

```{r}
expr <- expression((4/3)*pi*(d/2)^3)
derivative <- deriv(expr, 'd')
derivative
```

```{r}
library(Deriv)

# Define the function
f <- function(d) (4/3)*pi*(d/2)^3

# Compute the derivative
df <- Deriv(f)

# Print the derivative function
print(df)

# Evaluate the derivative at a point, e.g., x = 2
df(1)
```

```{r}
# (a) the x-axis
V = 4.189
pi * pracma::integral(function(x) { 1 }, 1, 2)
pi * pracma::integral(function(x) { (-x + 3)^2 }, 1, 2) - pi

# (b) y = 2
V = 2.094
pi - pi * pracma::integral(function(x) { (-x + 3 - 2)^2 }, 1, 2)

# (c) the y-axis
V = 4.189
# (same as x-axis here, after transformation)
pi * pracma::integral(function(x) { (-x + 3)^2 }, 1, 2) - pi

# (d) x = 1
V = 1.047
pi * pracma::integral(function(x) { (-x + 3 - 1)^2 }, 1, 2)
```

