---
title: "DATA605 Homework 15"
author: "Keith Colella"
date: "2023-12-11"
output: html_document
---

```{r, message = FALSE, warning = FALSE}
library(Ryacas)
library(Deriv)
```

## Question 1

Find the equation of the regression line for the given points. Round any final values to the nearest hundredth, if necessary.

$$( 5.6, 8.8 ), ( 6.3, 12.4 ), ( 7, 14.8 ), ( 7.7, 18.2 ), ( 8.4, 20.8 )$$

#### Response

The regression line is the best fit line, or the line that minimizes the sum of squared errors (SSE). Like any line, we can express it as a function $y = mx + b$, where $m$ is our slope and $b$ is our intercept. If we're using regression notation, we would express it as $y = \alpha + \beta x$. The SSE can be formulated as $SSE = \Sigma [y_i - \alpha + \beta x_{i})]^2$.

So, we can find the parameters $\alpha$ and $\beta$ that minimize the SSE by taking the partial derivatives of the SSE function with respect to $\alpha$ and $\beta$. We can then set the resulting partial derivatives equal to zero (representing the minimums of our SSE curves) and solve the system of equations.

Let's start with the partial derivatives of $SSE = f(\alpha,\beta) = \Sigma [y_i - \alpha + \beta x_{i})]^2$.

$f_{\alpha} = \frac{\partial f}{\partial \alpha} = -2 \sigma [y_i - \alpha + \beta x_{i})]$

$f_{\beta} = \frac{\partial f}{\partial \beta} = 2 \sigma [y_i - \alpha + \beta x_{i})]x_i$

We can then set both equal to zero and solve the system. After quite a bit of rearranging, we get the following.

$\alpha = \bar{y} + \beta \bar{x}$

$\beta = \frac{\bar{y} \Sigma x_i - \Sigma y_ix_i}{\Sigma x^2_i - \bar{x} \Sigma x_i}$

```{r}
beta_calc <- function(x, y) {
  n <- length(x)
  x_bar <- mean(x)
  y_bar <- mean(y)
  
  sum_xi <- sum(x)
  sum_yi_xi <- sum(y * x)
  sum_xi_squared <- sum(x^2)
  
  beta <- (y_bar * sum_xi - sum_yi_xi) / (sum_xi_squared - x_bar * sum_xi)
  return(beta)
}

x <- c(5.6, 6.3, 7, 7.7, 8.4)
y <- c(8.8, 12.4, 14.8, 18.2, 20.8)

beta <- beta_calc(x,y)

alpha_calc <- function(x, y, beta) {
  return(mean(y) + beta * mean(x))
}

alpha <- alpha_calc(x, y, beta)

cat(
  'Final regression line: y = ', 
  round(alpha,2), ' + ', 
  round(beta,2), 'x', 
  sep = ''
)
```

Now, let's check this against R's lm function, which calculates the same thing!

```{r}
lm <- lm(y~x)

lm_alpha <- lm$coefficients[1]
lm_beta <- lm$coefficients[2]

cat(
  'LM function regression line: y = ',
  round(lm_alpha,2), ' + ',
  round(lm_beta,2), 'x', 
  sep = ''
)
```

It's the same!

## Question 2

Find all local maxima, local minima, and saddle points for the function given below. Write your answer(s) in the form 
$(x, y, z)$. Separate multiple points with a comma.

$$f(x, y) = 24x - 6xy^2 - 8y^3$$

#### Response

Finding local minima/maxima requires the second derivative test, which involves (i) finding critical points, (ii) calculating the second partial derivatives and (iii) evaluating the discriminant. 

First we need the critical points, which are the first partial derivatives evaluated at 0. The first partial derivatives are:

$$f_X = \frac{\partial f}{\partial x} = 24 - 6y^2$$

$$f_y = \frac{\partial f}{\partial y} = -12xy - 6y^2$$

When evaluated at zero, they give us $y \pm -2$ and $x = -4$. So, our critical points are $(-4,2)$ and $(-4,-2)$.

Next, we need the second partial derivatives for this function, solved as follows:

$$f_{xx} = \frac{\partial ^2 f}{\partial x^2} = 0$$

$$f_{yy} = \frac{\partial ^2 f}{\partial y^2} = -12x - 48y$$

$$f_{xy} = \frac{\partial ^2 f}{\partial x \partial y} = -12y$$

The discriminant $D$ is defined as follows.

$$D = f_{xx}f_{yy} - (f_{xy})^2 = 0 \times (-12x-48y) - (12y)^2$$

Finally, let's evaluate at our critical points, $(-4,2)$ and $(-4,-2)$.

$$D_{(-4,2)} = 0 \times (-12(-4)-48(2)) - (12(2))^2 = -576$$
$$D_{(-4,-2)} = 0 \times (-12(-4)-48(2)) - (12(-2))^2 = -576$$

In both cases, the discriminant is < 0. Per the second derivative test definition, this indicates that both points are saddles, and the function has no local minima or maxima.

## Question 3

A grocery store sells two brands of a product, the "house" brand and a "name" brand. The manager estimates that if she sells the "house" brand for x dollars and the "name" brand for y dollars, she will be able to sell $81 - 21x + 17y$ units of the "house" brand and $40 + 11x - 23y$ units of the "name" brand.

**Step 1:** Find the revenue function $R (x, y)$.

**Step 2:** What is the revenue if she sells the "house" brand for \$2.30 and the "name" brand for \$4.10?

#### Response

Revenue for each product would be determined by the number of units sold multiplied by the price. We are given all of these values in terms of $x$ and $y$, so we can write this as a function.

$$R(x,y) = x(81 - 21x + 17y) + y(40 + 11x - 23y)$$

To evaluate this function with the provided values for $x$ and $y$, I'll write a quick functions.

```{r}
revenue <- function(x,y) {
  revenue <- x * (81 - 21*x + 17*y) + y * (40 + 11*x - 23*y)
  return(revenue)
}

rev <- revenue(x = 2.30, y = 4.10)

cat('Projected revenue: $', round(rev,2), sep = '')
```

## Question 4

A company has a plant in Los Angeles and a plant in Denver. The firm is committed to produce a total of 96 units of a product each week. The total weekly cost is given by 

$$C(x, y) = \frac{1}{6}x^2 + \frac{1}{6}y^2 + 7x + 25y + 700$$

where $x$ is the number of units produced in Los Angeles and $y$ is the number of units produced in Denver. How many units should be produced in each plant to minimize the total weekly cost?

#### Response

We can use a similar approach as we did when finding the SSE. First, we will find  

## Question 5

Evaluate the double integral on the given region.

$\iint (e^{8x + 3y}) dA$; $R:2 \leq x \leq 4$ and $2 \leq y \leq 4$
 
Write your answer in exact form without decimals.

#### Response

TBD