---
title: "DATA605 Homework 7"
author: "Keith Colella"
date: "2023-10-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

Let $X_1$, $X_2$, . . . , $X_n$ be n mutually independent random variables, each of which is uniformly distributed on the integers from 1 to $k$. Let $Y$ denote the minimum of the $X_i$’s. Find the distribution of $Y$.

### Response

Our goal is to 

# Question 2

Your organization owns a copier (future lawyers, etc.) or MRI (future doctors). This machine has a manufacturer’s expected lifetime of 10 years. This means that we expect one failure every ten years. (Include the probability statements and R Code for each part.)

## Part A

What is the probability that the machine will fail after 8 years? Provide also the expected value and standard deviation. Model as a geometric.
*Hint: the probability is equivalent to not failing during the first 8 years.*

### Response

```{r}
# Parameters
p <- 1/10
q <- 1 - p

# Calculations
prob_fail_after_8 <- (q^8) * p
prob_fail_after_8 <- (q^8) * p
expected_value <- 1/p
standard_deviation <- sqrt(q/p^2)

# Results
cat("Probability of failure after 8 years:", prob_fail_after_8, "\n")
cat("Expected value:", expected_value, "\n")
cat("Standard deviation:", standard_deviation, "\n")
```

```{r}
prob_fail_after_8 <- pgeom(q, p, lower.tail = FALSE)
cat("Probability of failure after 8 years:", prob_fail_after_8, "\n")
```


## Part B

What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as an exponential.

### Response

TBD

## Part C

What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as a binomial (*Hint: 0 success in 8 years*).

### Response

TBD

## Part D

What is the probability that the machine will fail after 8 years?. Provide also the expected value and standard deviation. Model as a Poisson.

### Response

TBHD