---
title: "D605 HW3 - Keith Colella"
output: html_document
date: "2023-09-11"
---

```{r setup}
library(pracma)
```

# Problem Set 1

### Problem 1

What is the rank of matrix A? 
$$ A = 
\begin{bmatrix} 
1 & 2 & 3 & 4\\
-1 & 0 & 1 & 3\\
0 & 1 & -2 & 1\\
5 & 4 & -2 & -3
\end{bmatrix} $$

#### Response

```{r}
A = matrix(c(
   1, 2, 3, 4,
  -1, 0, 1, 3,
   0, 1,-2, 1,
   5, 4,-2,-3
), nrow = 4, byrow = TRUE)

A
```

Manually reduce to rref

```{r}

```

Check against pracma

```{r}
rref(A)
```

A has a rank of 4. We can also that it is full rank.

### Problem 2

Given an $mxn$ matrix where $m > n$, what is the maximum rank? What is the minimum rank, assuming that the matrix is non-zero?

#### Response

The maximum rank is $n$. In general, the rank of a matrix cannot exceded the number of rows or columns. So, if we have a matrix where $m > n$, then the number of columns, i.e. $n$, is our limiting factor. In other words, the rank cannot exceed $n$.

We can demonstrate this with some examples. The rank of a matrix can be determined as the count of the number of non-zero rows in the row-reduced echelon form of that matrix. If we row reduce matrices with more rows than columns (i.e. $m > n$), we'll find that the number of non-zero rows never exceeds $n$.

```{r}
determine_rank <- function(input_matrix) {
  reduced_matrix <- rref(input_matrix)
  nz_rows <- 0
  for (row in 1:dim(reduced_matrix)[1]) {
    ifelse(
      sum(reduced_matrix[row,]) != 0,
      nz_rows <- nz_rows + 1, 0
    )
  }
  return(nz_rows)
}

M <- round(matrix(runif(20,-10,10), nrow = 5),0)
cat('Dimensions: ',dim(M),'\nNon-zero rows: ', determine_rank(M),'\n\n')

M <- round(matrix(runif(18,-10,10), nrow = 6),0)
cat('Dimensions: ',dim(M),'\nNon-zero rows: ', determine_rank(M),'\n\n')

M <- round(matrix(runif(30,-10,10), nrow = 10),0)
cat('Dimensions: ',dim(M),'\nNon-zero rows: ', determine_rank(M),'\n\n')

M <- round(matrix(runif(28,-10,10), nrow = 7),0)
cat('Dimensions: ',dim(M),'\nNon-zero rows: ', determine_rank(M),'\n\n')

M <- round(matrix(runif(30,-10,10), nrow = 6),0)
cat('Dimensions: ',dim(M),'\nNon-zero rows: ', determine_rank(M),'\n\n')
```

The minimum rank of such a matrix is 1 (assuming it is non-zero). Again, the rank of a matrix can be counted by the number of non-zero rows in the matrix's reduced form. So long as an matrix has one non-zero element in it, the reduced form will have at least one non-zero row. And in that case, the rank will be at least one.

```{r}

```


### Problem 3

What is the rank of matrix B? 
$$ B = 
\begin{bmatrix} 
1 & 2 & 1\\
3 & 6 & 3\\
2 & 4 & 2\\
\end{bmatrix} $$

#### Response

```{r}
B = matrix(c(
   1, 2, 1,
   3, 6, 3,
   2, 3, 2
), nrow = 3, byrow = TRUE)

B
```

As above, we can convert this matrix to reduced-row echelon form, then count the number of non-zero rows.

```{r}
rref(B)
```

B has a rank of 2.

# Problem Set 2

Compute the eigenvalues and eigenvectors of the matrix A. You’ll need to show your work. You’ll need to write out the characteristic polynomial and show your solution.

$$ A = 
\begin{bmatrix} 
1 & 2 & 3\\
0 & 4 & 5\\
0 & 0 & 6\\
\end{bmatrix} $$

#### Response

```{r}
A = matrix(c(
   1, 2, 3,
   0, 4, 5,
   0, 0, 6
), nrow = 3, byrow = TRUE)

A
```

Use the characteristic polynomial of $A$: $p_{A}(x) = det(A-\lambda I)$. We can expand this further.

$$ \begin{vmatrix}
  1 - \lambda & 2 & 3 \\
  0 & 4 - \lambda & 5 \\ 
  0 & 0 & 6 - \lambda
\end{vmatrix} $$

We can then break this into a series of 2x2 determinants multiplied by the respective scalars.

$$ 1-\lambda \begin{vmatrix}4-\lambda & 5 \\ 0 & 6-\lambda \end{vmatrix} + 
  2 \begin{vmatrix}0 & 5 \\ 0 & 6-\lambda \end{vmatrix} +
  3 \begin{vmatrix}0 & 4-\lambda \\ 0 & 0 \end{vmatrix} $$

We can apply our $ad - bc$ rule for calculating 2x2 determinants to obtain the following.

$$ (1-\lambda)[(4-\lambda)(6-\lambda)-5(0)] + 
  2[0(6-\lambda)-5(0)] + 
  3[0(0)-(4-\lambda)(0)] \\
= (1-\lambda)(24-4\lambda-6\lambda-\lambda^2-0) +
  2(0-0) + 3(0-0) \\
= (1-\lambda)(\lambda^2-10\lambda+24)+0+0 \\
= (1-\lambda)(\lambda-6)(\lambda-4) \\
= -(\lambda-1)(\lambda-6)(\lambda-4) $$

This indicates that our roots are $6$, $4$, and $1$, which corresponds to our three eigenvalues. 

We can use these eigenvalues to find the corresponding eigenspaces (and from those, eigenvectors). We'll begin by substituting each $\lambda$ back into the expanded characteristic polynomial (i.e. $A-\lambda I_{3}$). Starting with $\lambda = 6$, we get the following.

$$ \begin{bmatrix}
  1 - 6 & 2 & 3 \\
  0 & 4 - 6 & 5 \\ 
  0 & 0 & 6 - 6
\end{bmatrix} \\
= \begin{bmatrix}
  -5 & 2 & 3 \\
  0 & -2 & 5 \\ 
  0 & 0 & 0
\end{bmatrix} $$

We can then apply row operations to bring the matrix to reduced-row echelon form.

$$ \begin{bmatrix}
  -5 & 2 & 3 \\
  0 & -2 & 5 \\ 
  0 & 0 & 0
\end{bmatrix} 
\xrightarrow{R_{2}+R_{1}}
\begin{bmatrix}
  -5 & 0 & 8 \\
  0 & -2 & 5 \\ 
  0 & 0 & 0
\end{bmatrix} 
\xrightarrow{-1/2R_{2}}
\begin{bmatrix}
  -5 & 0 & 8 \\
  0 & -2 & -5/2 \\ 
  0 & 0 & 0
\end{bmatrix} 
\xrightarrow{-1/5R_{1}}
\begin{bmatrix}
  1 & 0 & -8/5 \\
  0 & -2 & -5/2 \\ 
  0 & 0 & 0
\end{bmatrix} $$

We know that the eigenspace is a null space of our characteristic polynomial, i.e. $\mathcal{E}_{A}(3) =  \mathcal{N}(A-6I_{3})$. We used the reduced form of this matrix to express the eigenspace as the span of a basis. We form a linear combination ($\times10$) to get to whole numbers.

$$ \mathcal{N}(A-6I_{3}) = 
\begin{Bmatrix} 8/5 \\ 5/2 \\ 1 \end{Bmatrix} = 
\begin{Bmatrix} 16 \\ 25 \\ 10 \end{Bmatrix} $$

We can apply the same procedure with $\lambda = 4$.

$$ \begin{bmatrix}
  1 - 4 & 2 & 3 \\
  0 & 4 - 4 & 5 \\ 
  0 & 0 & 6 - 4
\end{bmatrix}
= \begin{bmatrix}
  -3 & 2 & 3 \\
  0 & 0 & 5 \\ 
  0 & 0 & 2
\end{bmatrix} 
\xrightarrow{RREF}
\begin{bmatrix}
  1 & -2/3 & 0 \\
  0 & 0 & 1 \\ 
  0 & 0 & 0
\end{bmatrix} \\~\\
\mathcal{N}(A-3I_{3}) = 
\begin{Bmatrix} 2/3 \\ 1 \\ 0 \end{Bmatrix} =
\begin{Bmatrix} 2 \\ 3 \\ 0 \end{Bmatrix} $$

And again with $\lambda = 1$.

$$ \begin{bmatrix}
  1 - 1 & 2 & 3 \\
  0 & 4 - 1 & 5 \\ 
  0 & 0 & 6 - 1
\end{bmatrix}
= \begin{bmatrix}
  0 & 2 & 3 \\
  0 & 3 & 5 \\ 
  0 & 0 & 5
\end{bmatrix} 
\xrightarrow{RREF}
\begin{bmatrix}
  0 & 1 & 0 \\
  0 & 0 & 1 \\ 
  0 & 0 & 0
\end{bmatrix} \\~\\
\mathcal{N}(A-3I_{3}) = 
\begin{Bmatrix} 1 \\ 0 \\ 0 \end{Bmatrix} $$

We can now validate that our results observe the key properties of eigenvalues and eigenvectors. Namely, we'll confirm that $Av=\lambda v$ for all combinations of values and vectors.

```{r}
eigenvalues = c(6,4,1)
eigenvectors = list(
  matrix(c(16,25,10), nrow = 3),
  matrix(c(2,3,0), nrow = 3),
  matrix(c(1,0,0), nrow = 3)
)

for (i in 1:3) {
  cat('For λ =', eigenvalues[[i]], '\n', 'Av\n')
  print(A %*% eigenvectors[[i]])
  cat('λv\n')
  print(eigenvalues[i] * eigenvectors[[i]])
  cat('\n')
}
```

They match!

Finally, we'll check our week against R's native `eigen` function.

```{r}
eigen_A = eigen(A)

eigen_A
```

We can clean up the eigenvectors a bit...

```{r}
eigen_A[[2]][,1] * 10 / eigen_A[[2]][3,1]

eigen_A[[2]][,2] * 2 / eigen_A[[2]][1,2]

eigen_A[[2]][,3]
```

They match!