---
title: "Introduction to data"
author: ""
output:
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(openintro)
```

Some define statistics as the field that focuses on turning information into knowledge. The first step in that process is to summarize and describe the raw information -- the data. In this lab we explore flights, specifically a random sample of domestic flights that departed from the three major New York City airports in 2013. We will generate simple graphical and numerical summaries of data on these flights and explore delay times. Since this is a large data set, along the way you'll also learn the indispensable skills of data processing and subsetting.


## Getting started

### Load packages

In this lab, we will explore and visualize the data using the **tidyverse** suite of packages. The data can be found in the companion package for OpenIntro labs, **openintro**.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
```

### The data

The [Bureau of Transportation Statistics](http://www.rita.dot.gov/bts/about/) (BTS) is a statistical agency that is a part of the Research and Innovative Technology Administration (RITA). As its name implies, BTS collects and makes transportation data available, such as the flights data we will be working with in this lab.

First, we'll view the `nycflights` data frame. Type the following in your console to load the data:

```{r load-data}
data(nycflights)
```

The data set `nycflights` that shows up in your workspace is a *data matrix*, with each row representing an *observation* and each column representing a *variable*. R calls this data format a **data frame**, which is a term that will be used throughout the labs. For this data set, each *observation* is a single flight.

To view the names of the variables, type the command

```{r names}
names(nycflights)
```

This returns the names of the variables in this data frame. The **codebook** (description of the variables) can be accessed by pulling up the help file:

```{r help, eval=FALSE}
?nycflights
```

One of the variables refers to the carrier (i.e. airline) of the flight, which is coded according to the following system.

- `carrier`: Two letter carrier abbreviation.
    + `9E`:           Endeavor Air Inc.
    + `AA`:      American Airlines Inc.
    + `AS`:        Alaska Airlines Inc.
    + `B6`:             JetBlue Airways
    + `DL`:        Delta Air Lines Inc.
    + `EV`:    ExpressJet Airlines Inc.
    + `F9`:      Frontier Airlines Inc.
    + `FL`: AirTran Airways Corporation
    + `HA`:      Hawaiian Airlines Inc.
    + `MQ`:                   Envoy Air
    + `OO`:       SkyWest Airlines Inc.
    + `UA`:       United Air Lines Inc.
    + `US`:             US Airways Inc.
    + `VX`:              Virgin America
    + `WN`:      Southwest Airlines Co.
    + `YV`:          Mesa Airlines Inc.


Remember that you can use `glimpse` to take a quick peek at your data to understand its contents better.

```{r glimpse}
glimpse(nycflights)
```

The `nycflights` data frame is a massive trove of information. Let's think about some questions we might want to answer with these data:

- How delayed were flights that were headed to Los Angeles?
- How do departure delays vary by month?
- Which of the three major NYC airports has the best on time percentage for departing flights?


## Analysis


### Departure delays

Let's start by examing the distribution of departure delays of all flights with a histogram.

```{r hist-dep-delay}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram()
```

This function says to plot the `dep_delay` variable from the `nycflights` data frame on the x-axis. It also defines a `geom` (short for geometric object), which describes the type of plot you will produce. 

Histograms are generally a very good way to see the shape of a single distribution of numerical data, but that shape can change depending on how the data is split between the different bins. You can easily define the binwidth you want to use:

```{r hist-dep-delay-bins}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 15)
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 150)
```

1.  Look carefully at these three histograms. How do they compare? Are features revealed in one that are obscured in another?
 
#### Start Response 1

The plot with the narrowest bin widths provides a more granular view of the distribution, as expected. This granularity reveals some unique insights. Specifically, we see a significant number of observations to the left of the median bin, and these observations are "hidden" (or subsumed by the median bin) in the other two plots. Similarly the first bins in the first and third plots span negative and positive numbers, so we're unable to distinguish between flights that left early/on-time and those with minor delays.

Narrower bins tend to reveal these more granular details, but sometimes at a cost. Specifically, if we have subsequent bins with no observations, the resulting histogram will be somewhat distorted or "bumpy", and may give the impression of a multi-modal distribution. This contrasts with more appropriately sized bins, which, while obscuring some granularity, tend to provide a smoother view that might better support inference.

If we take the flights data as an example, we do not see this downside materialize. The dataset if sufficiently robust that, even with a bin width of 1, we have observations and in every bucket and a relatively smooth distribution.

```{r Response-1}
ggplot(data = nycflights, aes(x = dep_delay)) +
  geom_histogram(binwidth = 1)
```

#### End Response 1

If you want to visualize only on delays of flights headed to Los Angeles, you need to first `filter` the data for flights with that destination (`dest == "LAX"`) and then make a histogram of the departure delays of only those flights.

```{r lax-flights-hist}
lax_flights <- nycflights %>%
  filter(dest == "LAX")
ggplot(data = lax_flights, aes(x = dep_delay)) +
  geom_histogram()
```

Let's decipher these two commands (OK, so it might look like four lines, but the first two physical lines of code are actually part of the same command. It's common to add a break to a new line after `%>%` to help readability).

- Command 1: Take the `nycflights` data frame, `filter` for flights headed to LAX, and save the result as a new data frame called `lax_flights`.
    + `==` means "if it's equal to".
    + `LAX` is in quotation marks since it is a character string.
- Command 2: Basically the same `ggplot` call from earlier for making a histogram, except that it uses the smaller data frame for flights headed to LAX instead of all flights.

<div id="boxedtext">
**Logical operators: ** Filtering for certain observations (e.g. flights from a particular airport) is often of interest in data frames where we might want to examine observations with certain characteristics separately from the rest of the data. To do so, you can use the `filter` function and a series of **logical operators**. The most commonly used logical operators for data analysis are as follows:

- `==` means "equal to"
- `!=` means "not equal to"
- `>` or `<` means "greater than" or "less than"
- `>=` or `<=` means "greater than or equal to" or "less than or equal to"
</div>

You can also obtain numerical summaries for these flights:

```{r lax-flights-summ}
lax_flights %>%
  summarise(mean_dd   = mean(dep_delay), 
            median_dd = median(dep_delay), 
            n         = n())
```

Note that in the `summarise` function you created a list of three different numerical summaries that you were interested in. The names of these elements are user defined, like `mean_dd`, `median_dd`, `n`, and you can customize these names as you like (just don't use spaces in your names). Calculating these summary statistics also requires that you know the function calls. Note that `n()` reports the sample size.

<div id="boxedtext">
**Summary statistics: ** Some useful function calls for summary statistics for a single numerical variable are as follows:

- `mean`
- `median`
- `sd`
- `var`
- `IQR`
- `min`
- `max`

Note that each of these functions takes a single vector as an argument and returns a single value. 
</div>

You can also filter based on multiple criteria. Suppose you are interested in flights headed to San Francisco (SFO) in February:

```{r sfo-feb-flights}
sfo_feb_flights <- nycflights %>%
  filter(dest == "SFO", month == 2)
```

Note that you can separate the conditions using commas if you want flights that are both headed to SFO **and** in February. If you are interested in either flights headed to SFO **or** in February, you can use the `|` instead of the comma.

2.  Create a new data frame that includes flights headed to SFO in February, 
    and save this data frame as `sfo_feb_flights`. How many flights 
    meet these criteria?
    
#### Start Response 2
```{r Response-2}
sfo_feb_flights <- nycflights %>%
  filter(dest == "SFO", month == 2)

summarize(sfo_feb_flights, n = n())
```
There are 68 flights in the dataset that went to SFO in February.

#### End Response 2
    
3.  Describe the distribution of the **arrival** delays of these flights using a 
    histogram and appropriate summary statistics. **Hint:** The summary 
    statistics you use should depend on the shape of the distribution.

#### Start Response 3
```{r Response-3}
ggplot(sfo_feb_flights, aes(x = arr_delay)) +
  geom_histogram(aes(y = ..density..), binwidth = 10) +
  stat_function(fun = dnorm,
                args = list(mean = mean(sfo_feb_flights$arr_delay), 
                            sd = sd(sfo_feb_flights$arr_delay)
                            )
                )

sfo_feb_flights %>%
  summarize(mean_ar = mean(arr_delay),
            median_ar = median(arr_delay),
            sd_ar = sd(arr_delay)
            )
```
Both the mean and median indicate that most February flights to SFO arrived early. The are some outliers in the right tail that appear to skew the mean, so the median of -11 minutes appears to be the better measure of centrality.

Plotting a normal shows a pretty weak fit to the distribution. Again, the outliers appear to skew the summary stats, giving the indication of greater spread / heavier tails.

#### End Response 3
    
Another useful technique is quickly calculating summary statistics for various groups in your data frame. For example, we can modify the above command using the `group_by` function to get the same summary stats for each origin airport:

```{r summary-custom-list-origin}
sfo_feb_flights %>%
  group_by(origin) %>%
  summarise(median_dd = median(dep_delay), iqr_dd = IQR(dep_delay), n_flights = n())
```

Here, we first grouped the data by `origin` and then calculated the summary statistics.

4.  Calculate the median and interquartile range for `arr_delay`s of flights in
    in the `sfo_feb_flights` data frame, grouped by carrier. Which carrier
    has the most variable arrival delays?
    
#### Start Response 4
```{r Response-4}
sfo_feb_flights %>%
  group_by(carrier) %>%
  summarize(median_ad = median(arr_delay),
            iqr_ad = IQR(arr_delay),
            sq_ad = sd(arr_delay)
            )
```
Based on IQR, Delta and United appear tied in terms of variability. Introducing standard deviation, however, indicates that United actually has greater variability in arrival times. Virgin also appears to have a higher SD than Delta, although its IQR is narrower.

#### End Response 4

### Departure delays by month

Which month would you expect to have the highest average delay departing from an NYC airport?

Let's think about how you could answer this question:

- First, calculate monthly averages for departure delays. With the new language you are learning, you could
    + `group_by` months, then
    + `summarise` mean departure delays.
- Then, you could to `arrange` these average delays in `desc`ending order

```{r mean-dep-delay-months}
nycflights %>%
  group_by(month) %>%
  summarise(mean_dd = mean(dep_delay)) %>%
  arrange(desc(mean_dd))
```

5.  Suppose you really dislike departure delays and you want to schedule 
    your travel in a month that minimizes your potential departure delay leaving 
    NYC. One option is to choose the month with the lowest mean departure delay.
    Another option is to choose the month with the lowest median departure delay. 
    What are the pros and cons of these two choices?

#### Start Response 5
```{r Response-5}
nycflights %>%
  group_by(month) %>%
  summarize(mean_dd = mean(dep_delay),
            median_dd = median(dep_delay),
            min_dd = min(dep_delay),
            max_dd = max(dep_delay)
            ) %>%
  mutate(rank_mean = min_rank(mean_dd),
         rank_med = min_rank(median_dd),
         avg_rank = (rank_mean + rank_med) / 2
         ) %>%
  arrange(avg_rank)

ggplot(nycflights, aes(x = dep_delay, group = month, color = month)) +
  geom_density()
```
The skewed distribution of departure delays again drives significant differences between mean and median measures. The mean appears higher in all cases because of the large number of very long delays. This concentration of observations in the right tail can be seen in the density plot, as well as the disparity between min and max values (despite the data being largely centered on zero).

In terms of the best month to travel, both measures point to October. The pro of choosing the median is that we eliminate some of the "noise" introduced by flights with very long delays. The con, however, is the essentially the same. By eliminating the impact of those mega delays, we might also be eliminating real "signal" in the data. So, by using the median, we provide the best chance of an early departure, but we ignore the potential of an idiosyncratic mega delay.

Finally, I'll propose an alterantive measure. Rather than identifying the month with the lowest mean/median delay, we can search for the month with the volume of delays overall, measured as a proportion of total flights. Ultimately, however, the answer is quite similar (as one might expect). September has the lowest proportion of delayed flights, followed closely by October.
```{r}
nycflights %>%
  group_by(month) %>%
  mutate(delay = case_when(dep_delay > 0 ~ 1,
                           dep_delay <= 0 ~ 0
                           )
         ) %>%
  summarize(total = n(),
            delays = sum(delay),
            delay_prop = (delays / total)
            ) %>%
  arrange(delay_prop)
```
#### End Response 5

### On time departure rate for NYC airports

Suppose you will be flying out of NYC and want to know which of the three major NYC airports has the best on time departure rate of departing flights. Also supposed that for you, a flight that is delayed for less than 5 minutes is basically "on time."" You consider any flight delayed for 5 minutes of more to be "delayed".

In order to determine which airport has the best on time departure rate, you can 

- first classify each flight as "on time" or "delayed",
- then group flights by origin airport,
- then calculate on time departure rates for each origin airport,
- and finally arrange the airports in descending order for on time departure percentage.

Let's start with classifying each flight as "on time" or "delayed" by creating a new variable with the `mutate` function.

```{r dep-type}
nycflights <- nycflights %>%
  mutate(dep_type = ifelse(dep_delay < 5, "on time", "delayed"))
```

The first argument in the `mutate` function is the name of the new variable we want to create, in this case `dep_type`. Then if `dep_delay < 5`, we classify the flight as `"on time"` and `"delayed"` if not, i.e. if the flight is delayed for 5 or more minutes.

Note that we are also overwriting the `nycflights` data frame with the new version of this data frame that includes the new `dep_type` variable.

We can handle all of the remaining steps in one code chunk:

```{r ot-dep-rate}
nycflights %>%
  group_by(origin) %>%
  summarise(ot_dep_rate = sum(dep_type == "on time") / n()) %>%
  arrange(desc(ot_dep_rate))
```

6.  If you were selecting an airport simply based on on time departure percentage, which NYC airport would you choose to fly out of?

#### Start Response 6
LGA has the highest percentage of on time departures (including the 5min allowance), so if I would chooese LGA. The graph below helps to directly compare on-time rate across the three airports.
```{r Response-6}
nycflights %>%
  group_by(origin) %>%
  summarize(ot_dep_rate = sum(dep_type == 'on time') / n(),
            dep_type = dep_type) %>%
  ggplot(aes(x = origin, y = ot_dep_rate, fill = dep_type)) +
    geom_bar(position = "fill", stat='identity')
```
#### End Response 6

You can also visualize the distribution of on on time departure rate across the three airports using a segmented bar plot.

```{r viz-origin-dep-type}
ggplot(data = nycflights, aes(x = origin, fill = dep_type)) +
  geom_bar()
```

* * *

## More Practice

7.  Mutate the data frame so that it includes a new variable that contains the 
    average speed, `avg_speed` traveled by the plane for each flight (in mph).
    **Hint:** Average speed can be calculated as distance divided by
    number of hours of travel, and note that `air_time` is given in minutes.

#### Start Response 7
TBD
```{r Response-7}
nycflights <- nycflights %>%
  mutate(avg_speed = distance / air_time * 60)

nycflights[1:10,] %>%
  select('carrier', 'tailnum', 'flight', 'dest', 'air_time', 'distance', 'avg_speed') %>%
  arrange(desc(avg_speed))
```
#### End Response 7

8.  Make a scatterplot of `avg_speed` vs. `distance`. Describe the relationship
    between average speed and distance.
    **Hint:** Use `geom_point()`.

#### Start Response 9
TBD
```{r Response-8}
ggplot(nycflights, aes(x = avg_speed, y = distance, color = dep_type)) +
  geom_point()
```
#### End Response 8

9.  Replicate the following plot. **Hint:** The data frame plotted only
    contains flights from American Airlines, Delta Airlines, and United
    Airlines, and the points are `color`ed by `carrier`. Once you replicate
    the plot, determine (roughly) what the cutoff point is for departure
    delays where you can still expect to get to your destination on time.

```{r plot-to-replicate, echo=FALSE, fig.show="asis", fig.width=7, fig.height=4}
dl_aa_ua <- nycflights %>%
  filter(carrier == "AA" | carrier == "DL" | carrier == "UA")
ggplot(data = dl_aa_ua, aes(x = dep_delay, y = arr_delay, color = carrier)) +
  geom_point()
```

#### Start Response 9
TBD
```{r Response-9}
dl_aa_ua <- nycflights %>%
  filter(carrier == "AA" | carrier == "DL" | carrier == "UA")

ggplot(data = dl_aa_ua, aes(x = dep_delay, y = arr_delay, color = carrier)) +
  geom_point()
```

~63 minutes
```{r}
dl_aa_ua %>%
  filter(arr_delay <= 0) %>%
  ggplot(aes(x = dep_delay, y = arr_delay, color = carrier)) +
  geom_point()
```

#### End Response 9