---
title: "Lab 8 - Linear Regressions"
author: "Keith Colella"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../D606/output"
    )
  })
---

## Setup

Config

```{r setup, message = FALSE}
library(tidyverse)
library(openintro)
# library(infer)
# library(cowplot)
# library(kableExtra)
set.seed(789)
```

Data

```{r}
data('hfi', package='openintro')
```

## Exercise 1

#### Question

What are the dimensions of the dataset?

#### Response

```{r}
nrow(hfi)
ncol(hfi)
```

The dataset has 123 variables (columns) with 1458 observations (rows).

---

## Exercise 2

#### Question

What type of plot would you use to display the relationship between the personal freedom score, `pf_score`, and one of the other numerical variables? Plot this relationship using the variable `pf_expression_control` as the predictor. Does the relationship look linear? If you knew a country’s `pf_expression_control`, or its score out of 10, with 0 being the most, of political pressures and controls on media content, would you be comfortable using a linear model to predict the personal freedom score?

#### Response

```{r}
ggplot(hfi, aes(pf_score, pf_expression_control)) +
  geom_point()
```

A scatter plot is a helpful way to explore potential linear relationships between two variables. Typically, the response variable takes the x axis, and the predictor takes the y axis. There does appear to be a reasonably linear, positive relationship between the personal freedom score and the expression control score. So, a linear model may be appropriate to predict `pf_score` based on `pf_expression_control`.

---

## Exercise 3

#### Question

Looking at your plot from the previous exercise, describe the relationship between these two variables. Make sure to discuss the form, direction, and strength of the relationship as well as any unusual observations.

```{r}
hfi %>%
  summarise(cor(pf_expression_control, pf_score, use = "complete.obs"))
```


#### Response

The relationship appears positive. That is, as `pf_expression_control` increases, so does `pf_score`. The relationship appears relatively strong, as indicated by the correlation coefficient of ~0.80. There are, however,   

---

## Exercise 4

#### Question

Using plot_ss, choose a line that does a good job of minimizing the sum of squares. Run the function several times. What was the smallest sum of squares that you got? How does it compare to your neighbors?

#### Response

```{r}

```

TBD.

---

## Exercise 5

#### Question

Fit a new model that uses pf_expression_control to predict hf_score, or the total human freedom score. Using the estimates from the R output, write the equation of the regression line. What does the slope tell us in the context of the relationship between human freedom and the amount of political pressure on media content?

#### Response

TBD

---

## Exercise 6

#### Question

If someone saw the least squares regression line and not the actual data, how would they predict a country’s personal freedom school for one with a 6.7 rating for pf_expression_control? Is this an overestimate or an underestimate, and by how much? In other words, what is the residual for this prediction?

#### Response

```{r}

```

TBD

---

## Exercise 7

#### Question

Is there any apparent pattern in the residuals plot? What does this indicate about the linearity of the relationship between the two variables?

#### Response

```{r, warning = FALSE}

```

TBD

---

## Exercise 8

#### Question

Based on the histogram and the normal probability plot, does the nearly normal residuals condition appear to be met?

#### Response

```{r}

```

TBD

---

## Exercise 9

#### Question

Based on the residuals vs. fitted plot, does the constant variability condition appear to be met?

#### Response

```{r}

```

TBD

---

## Exercise 10

#### Question

Choose another freedom variable and a variable you think would strongly correlate with it.. Produce a scatterplot of the two variables and fit a linear model. At a glance, does there seem to be a linear relationship?

#### Response

```{r}

```

TBD

---

## Exercise 11

#### Question

How does this relationship compare to the relationship between pf_expression_control and pf_score? Use the R2
 values from the two model summaries to compare. Does your independent variable seem to predict your dependent one better? Why or why not?

#### Response

```{r}

```

TBD

---

## Exercise 12

#### Question

What’s one freedom relationship you were most surprised about and why? Display the model diagnostics for the regression model analyzing this relationship.

#### Response

```{r}

```

TBD

---