---
title: "Lab 9 - Multiple Linear Regressions"
author: "Keith Colella"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../D606/output"
    )
  })
---

## Setup

Config

```{r setup, message = FALSE}
library(tidyverse)
library(openintro)
library(GGally)
```

Data

```{r}
data('evals', package='openintro')
glimpse(evals)
```

## Exercise 1

#### Question

Is this an observational study or an experiment? The original research question posed in the paper is whether beauty leads directly to the differences in course evaluations. Given the study design, is it possible to answer this question as it is phrased? If not, rephrase the question.

#### Response

The dataset has 123 variables (columns) with 1458 observations (rows).

---

## Exercise 2

#### Question

Describe the distribution of `score`. Is the distribution skewed? What does that tell you about how students rate courses? Is this what you expected to see? Why, or why not?

#### Response



---

## Exercise 3

#### Question

Excluding `score`, select two other variables and describe their relationship with each other using an appropriate visualization.

#### Response



---

## Exercise 4

#### Question

Replot the scatterplot, but this time use `geom_jitter` as your layer. What was misleading about the initial scatterplot?

```{r}

```


#### Response



---

## Exercise 5

#### Question

Let’s see if the apparent trend in the plot is something more than natural variation. Fit a linear model called `m_bty` to predict average professor score by average beauty rating. Write out the equation for the linear model and interpret the slope. Is average beauty score a statistically significant predictor? Does it appear to be a practically significant predictor?

#### Response



---

## Exercise 6

#### Question

Use residual plots to evaluate whether the conditions of least squares regression are reasonable. Provide plots and comments for each one (see the Simple Regression Lab for a reminder of how to make these).

#### Response



---

## Exercise 7

#### Question

P-values and parameter estimates should only be trusted if the conditions for the regression are reasonable. Verify that the conditions for this model are reasonable using diagnostic plots.

#### Response



---

## Exercise 8

#### Question

Is `bty_avg` still a significant predictor of score? Has the addition of `gender` to the model changed the parameter estimate for `bty_avg`?

#### Response



---

## Exercise 9

#### Question

What is the equation of the line corresponding to those with color pictures? (*Hint*: For those with color pictures, the parameter estimate is multiplied by 1.) For two professors who received the same beauty rating, which color picture tends to have the higher course evaluation score?

#### Response



---

## Exercise 10

#### Question

Create a new model called `m_bty_rank` with `gender` removed and `rank` added in. How does R appear to handle categorical variables that have more than two levels? Note that the rank variable has three levels: `teaching`, `tenure track`, `tenured`.

#### Response



---

## Exercise 11

#### Question

Which variable would you expect to have the highest p-value in this model? Why? *Hint*: Think about which variable would you expect to not have any association with the professor score.

#### Response



---

## Exercise 12

#### Question

Check your suspicions from the previous exercise. Include the model output in your response.

#### Response



---

## Exercise 13

#### Question

Interpret the coefficient associated with the ethnicity variable.

#### Response



---

## Exercise 14

#### Question

Drop the variable with the highest p-value and re-fit the model. Did the coefficients and significance of the other explanatory variables change? (One of the things that makes multiple regression interesting is that coefficient estimates depend on the other variables that are included in the model.) If not, what does this say about whether or not the dropped variable was collinear with the other explanatory variables?

#### Response



---

## Exercise 15

#### Question

Using backward-selection and p-value as the selection criterion, determine the best model. You do not need to show all steps in your answer, just the output for the final model. Also, write out the linear model for predicting score based on the final model you settle on.

#### Response



---

## Exercise 16

#### Question

Verify that the conditions for this model are reasonable using diagnostic plots.

#### Response



---

## Exercise 17

#### Question

The original paper describes how these data were gathered by taking a sample of professors from the University of Texas at Austin and including all courses that they have taught. Considering that each row represents a course, could this new information have an impact on any of the conditions of linear regression?

#### Response



---

## Exercise 18

#### Question

Based on your final model, describe the characteristics of a professor and course at University of Texas at Austin that would be associated with a high evaluation score.

#### Response



---

## Exercise 19

#### Question

Would you be comfortable generalizing your conclusions to apply to professors generally (at any university)? Why or why not?

#### Response



---