---
title: "Project DataExploration Candidates"
author: "Keith Colella"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../D606/output"
    )
  })
---

```{r setup, message=FALSE}
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
library(plotly)
library(scales)
```

## Election Results Data

MIT Election Lab
https://electionlab.mit.edu/data

U.S. House 1976–2020
https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/IG0UN2

```{r election-data}
elections <- read_csv('data/1976-2020-house.csv')

glimpse(elections)

elections <- filter(elections, stage == 'GEN')
```

Calculate wasted votes per candidate
```{r}
elections <- mutate(elections, yearStateDist = paste0(year, state_po, district))

elections <- elections %>% 
  group_by(yearStateDist) %>% 
  mutate(result = case_when(candidatevotes == max(candidatevotes) ~ 'winner',
                            candidatevotes != max(candidatevotes) ~ 'loser')) %>%
  ungroup()


elections <- elections %>%
  mutate(wastedVotes = case_when(result == 'winner' ~ (candidatevotes - (totalvotes %/% 2)),
                                 result == 'loser' ~ candidatevotes),
         wastedVotes = case_when(party == 'DEMOCRAT' ~ -wastedVotes,
                                 party == 'REPUBLICAN' ~ wastedVotes,
                                 TRUE ~ 0)
         )

elections %>%
  select(year, state_po, district, candidate, party, 
         candidatevotes, totalvotes, result, wastedVotes) %>%
  head(9)
```

Calculate efficiency gap per state per year
```{r}
eff_gap <- elections %>%
  group_by(year, state_po) %>%
  summarize(netWastedVotes = sum(wastedVotes),
            totalVotes = sum(totalvotes),
            eff_gap = abs(netWastedVotes / totalVotes)
            ) %>%
  arrange(year, desc(eff_gap))

eff_gap %>% 
  mutate(eff_gap = percent(eff_gap, 0.01)) %>%
  filter(year == 2018)
```


## Representative Ideology Data

https://www.govtrack.us/about/analysis#ideology
https://www.govtrack.us/data/analysis/by-congress/

116th Congress was elected in 2018 mid-terms
117th Congress was elected in 2020 presidentials
118th Congress was elected in 2022 mid-terms

Additional info on sources, including mapping of Congressional Districts
https://www.govtrack.us/about-our-data

Other Congress Bio Info
https://bioguide.congress.gov/
https://data.world/govtrack/us-congress-legislators

```{r rep-data}
house2018 <- read_delim('https://www.govtrack.us/data/analysis/by-congress/116/sponsorshipanalysis_h.txt',
                         delim = ',', col_names = TRUE)

glimpse(house2018)

ggplot(house2018, aes(x = ideology, y = leadership, color = party)) +
  geom_point() + 
  scale_color_manual(values = c('Democrat' = 'blue',
                                'Republican' = 'red',
                                'Libertarian' = 'pink',
                                'Independent' = 'purple'))
```

```{r, message=FALSE}
house <- data.frame(matrix(ncol = 12, nrow = 0))
colnames(house) <- c('ID', 'ideology', 'leadership', 'name', 'party', 
                     'description', 'introduced_bills', 'cosponsored_bills', 
                     'unique_cosponsors', 'total_cosponsors', 'year', 'congress')
year <- 1976
congress <- 95

for (i in 95:117) {
  link <- paste0('https://www.govtrack.us/data/analysis/by-congress/',
                 i,'/sponsorshipanalysis_h.txt')
  # delim <- if_else(i < 115, ', ', ',') 
  # ith_house <- read_delim(link, delim = delim, col_names = TRUE)
  ith_house <- read_delim(link, delim = ',', trim_ws = TRUE)
  ith_house['year'] = year
  ith_house['congress'] = congress
  colnames(ith_house) <- colnames(house)
  house <- rbind(house, ith_house)
  year <- year + 2
  congress <- congress + 1
}
```


## Members of Congress data

Primarily to link the two datasets.
https://theunitedstates.io/

```{r member-data, message=FALSE}
members_current <- read_csv('https://theunitedstates.io/congress-legislators/legislators-current.csv')
members_history <- read_csv('https://theunitedstates.io/congress-legislators/legislators-historical.csv')

members <- rbind(members_history, members_current)
```

## Combining datasets

```{r}
combo <- left_join(house, members[, c('govtrack_id', 'state', 'district')], 
                   by = c('ID' = 'govtrack_id'))

combo <- left_join(mutate(combo, party = tolower(party)),
                   elections  %>%
                     mutate(state = state_po, 
                            district = as.integer(district), 
                            party = tolower(party)) %>%
                     select(year, state, district, party, candidatevotes, totalvotes),
                   by = c('year', 'state', 'district', 'party'))

combo <- left_join(combo, 
                   eff_gap[, c('year', 'state_po', 'eff_gap')], 
                   by = c('year', 'state' = 'state_po'))

combo <- mutate(combo, percent_vote = candidatevotes / totalvotes)

head(combo,10)
```


```{r}
p1 <- combo %>%
  ggplot(aes(x = ideology, y = percent_vote, color = party)) + 
  geom_point(aes(frame = year)) + 
  scale_color_manual(values = c('democrat' = 'blue',
                                'republican' = 'red',
                                'libertarian' = 'pink',
                                'independent' = 'purple'))

ggplotly(p1)
```


```{r}
p2 <- combo %>%
  ggplot(aes(x = ideology, y = eff_gap, color = party)) +
  geom_point(aes(frame = year)) + 
  scale_color_manual(values = c('democrat' = 'blue',
                                'republican' = 'red',
                                'libertarian' = 'pink',
                                'independent' = 'purple'))

ggplotly(p2)
```



