---
title: "CUNY SPS MSDS - DATA607 - Week 3"
author: "Keith Colella"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../D607/output"
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
```

## Task 1

Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], 
provide code that identifies the majors that contain either "DATA" or "STATISTICS".

```{r task1}
## Read in dataset from 538's git
majors <- read.csv(paste0('https://raw.githubusercontent.com/fivethirtyeight/',
                          'data/master/college-majors/majors-list.csv'))

## Loop on each row and check majors column for either 'DATA' or 'STATISTICS', 
## then print all matches
for (i in 1:nrow(majors)) {
  if(grepl('DATA',majors[i,'Major']) | 
     grepl('STATISTICS',majors[i,'Major'])) {
    print(majors[i,'Major'])
  }
}
```

## Task 2

Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"  
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"  
[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", 
"chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

```{r task2}

## Create input string that needs to be converted
input <- (
'
[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"    
[13] "olive"        "salal berry"
'
)

## Locate all quotes, to be used for parsing out substrings
locations <- str_locate_all(input, '"')[[1]]

## Initialize variables for loop
i = 1
str <- 'c('

## Loop on input string to extract string between 
## 1st and 2nd quote, then 3rd and 4th...then (n-1)th and nth
while (i < length(locations[,1])) {
  result <- str_sub(input, locations[i]+1, locations[i+1]-1)
  str <- str_c(str, ' "', result, '",')
  i <- i + 2
}

## Remove final space and add closing parenthesis
str <- str_sub(str, 1, str_length(str)-1)
str <- str_c(str, ')')

## Print to confirm. Note the string appears with backslashes to escape the quotes 
print(str)

## Double check final string by evaluating
eval(parse(text = str))
```

## Task 3

Describe, in words, what these expressions will match:

1. (.)\1\1
2. "(.)(.)\\2\\1"
3. (..)\1
4. "(.).\\1.\\1"
5. "(.)(.)(.).*\\3\\2\\1"

### Expression 1
```{r task3.1}
## This expression will find any set of three repeated characters.
## However, we must escape the backslash when using it in a string.
str1 <- "Hellooo good friend." 
str_view(str1, "(.)\\1\\1")
```

### Expression 2
```{r task3.2}
## This regex will match an "abba" type sequence, in which
## a string contains a character, then two repeat characters,
## then the first character again.
str2 <- "I like ABBA's music." 
str_view(str2, "(.)(.)\\2\\1")
```

### Expression 3
```{r task3.3}
## This regex will match any four repeat characters. Again, we
## must escape the backslash to use it in a string.
str3 <- "But I reeeeally like their costumes." 
str_view(str3, "(..)\\1")
```

### Expression 4
```{r task3.4}
## This regex will match a series start any character, then any other
## character, followed by the first character again, then any other
## character, followed by the first character once again.
str4a <- "I use different vowels when I chuckle haheh." 
str_view(str4a, "(.).\\1.\\1")
## This also works when the repeat characters are spaces.
str4b <- "a b c d e f g h i j"
str_view(str4b, "(.).\\1.\\1")
```

### Expression 5
```{r task3.5}
## This regex will find any three unique characters, followed by
## any amount of other characters, following by those same first
## three characters in reverse order.
str5a <- "abc cba" 
str_view(str5a, "(.)(.)(.).*\\3\\2\\1")
str5b <- "abc                         cba" 
str_view(str5b, "(.)(.)(.).*\\3\\2\\1")
str5c <- "abc otherstuff cba" 
str_view(str5c, "(.)(.)(.).*\\3\\2\\1")
```

## Task 4

Construct regular expressions to match words that:

6. Start and end with the same character.
7. Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
8. Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)


### Expression 6
```{r task4.6}
str6 <- c("tacos", "emote", "helicopter")
str_view(str6, "(.).*\\1")
## This gives strange matches of partial words that
## fit the pattern, so we can add word boundaries to
## focus only on whole words.
str_view(str6, "\\b(.).*\\1\\b")
```

### Expression 7
```{r task4.7}
str7 <- c("church", "banana", "helicopter")
str_view(str7, "(..).*\\1")
## Again, this gives strange partial word matches. So, we
## can add word boundaries and wildcards to grab full words
## with any repeated pairs of letters, rather than partial words.
str_view(str7, "\\b.*(..).*\\1.*\\b")
```

### Expression 8
```{r task4.8}
str8 <- c("eleven", "banana", "amazinga", "helicopter")
str_view(str8, "\\b.*(.).*\\1.*\\1.*\\b")
```