---
title: "CUNY SPS MSDS - DATA607 - Week 3"
author: "Keith Colella"
date: "`r Sys.Date()`"
output: html_document
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../D607/output"
    )
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
```

## Task 1

Using the 173 majors listed in fivethirtyeight.com’s College Majors dataset [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/], provide code that identifies the majors that contain either "DATA" or "STATISTICS"

```{r task1}
majors <- read.csv(paste('https://raw.githubusercontent.com/fivethirtyeight/',
                         'data/master/college-majors/majors-list.csv',
                         sep = ''))

for (i in 1:nrow(majors)) {
  if(grepl('DATA',majors[i,'Major']) | 
     grepl('STATISTICS',majors[i,'Major'])) {
    print(majors[i,'Major'])
  }
}
```

## Task 2

Write code that transforms the data below:

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"
[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  
[9] "elderberry"   "lime"         "lychee"       "mulberry"    
[13] "olive"        "salal berry"

Into a format like this:

c("bell pepper", "bilberry", "blackberry", "blood orange", "blueberry", "cantaloupe", "chili pepper", "cloudberry", "elderberry", "lime", "lychee", "mulberry", "olive", "salal berry")

```{r task2}

```

## Task 3

Describe, in words, what these expressions will match:
(.)\1\1
(.)(.)\\2\\1
(..)\1
(.).\\1.\\1
(.)(.)(.).*\\3\\2\\1

```{r task3}

```

## Task 4

Construct regular expressions to match words that:
Start and end with the same character.
Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

```{r task4}

```