---
title: "Project 1 - Model Workflow"
author: "Keith Colella"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, message = FALSE}
library(tidyverse)
library(forecast)
library(urca)
```

Choice of variable / model parameters

```{r}
# knitting
show_code <- FALSE

# data
category <- 'S02'
variable <- 'Var03'
frequency <- 21

# outliers
outlier_removal <- TRUE

# missing values (method must be 'linear' or 'spline')
interpolate <- TRUE
missing_method <- 'linear' 

# skewness (method must be 'BoxCox' or 'log')
transform <- FALSE
trans_method <- 'BoxCox'
transform_diff <- FALSE
trans_method_diff <- 'BoxCox'

# seasonality
seasonal_periods <- c(5,21,252)

# arima
p <- 2
d <- 1
q <- 2
drift <- FALSE

# auto arima
approximation <- FALSE
stepwise <- FALSE

# seasonal arima
p_s <- 1
d_s <- 0
q_s <- 1
drift_s <- FALSE

# final model
final_method <- 'arima'
final_p <- 2
final_d <- 1
final_q <- 2
final_p_s <- 0
final_d_s <- 0
final_q_s <- 0
write_to_csv <- FALSE
```

Read in data. Convert to ts object and plot.

```{r, echo = show_code}
xl <- readxl::read_excel('data/DATA624_Project1_Data_Schema.xlsx',
                 sheet = category, skip = 2)

series <- xl[1:1622,variable]

series <- ts(deframe(series), frequency = frequency)

head(series)
autoplot(series)
```

Split training and test sets

```{r, echo = show_code}
train <- head(series, round(length(series) * 0.7))
test <- tail(series, round(length(series) * 0.3))

cat('Obs. in test set:',length(train),
    '\nObs. in test set:',length(test),
    '\nTotal observations:',length(series))
    
autoplot(train)
autoplot(test)
```

Split training and test sets with differenced series

```{r, echo = show_code}
train_diff <- diff(train)
test_diff <- diff(test)

autoplot(train_diff)
autoplot(test_diff)
```

Baseline forecasts - Naive, Average and ETS

```{r, echo = show_code}
naive_model <- naive(train, h = length(test), level = 0)
autoplot(naive_model) + 
  autolayer(test, alpha = 0.5)

mean_model <- meanf(train, h = length(test), level = 0)
autoplot(mean_model) + 
  autolayer(test, alpha = 0.5)

ets_model <- ets(train, na.action = 'na.interp')
ets_forecast <- forecast(ets_model, h = length(test), PI = FALSE)
autoplot(ets_forecast) +
  autolayer(test, alpha = 0.5)
```

Baseline forecasts on differenced series

```{r, echo = show_code}
naive_model_d <- naive(train_diff, h = length(test), level = 0)
naive_model_d_levels <- ts(cumsum(c(tail(train,1), naive_model_d$mean)),
                           frequency = frequency, start = time(test)[1])
autoplot(train) + 
  autolayer(naive_model_d_levels) + 
  autolayer(test, alpha = 0.5) + 
  scale_color_manual(values = c('blue','red'),
                     label = c('actual','prediction'))

mean_model_d <- meanf(train_diff, h = length(test), level = 0)
mean_model_d_levels <- ts(cumsum(c(tail(train,1), mean_model_d$mean)),
                          frequency = frequency, start = time(test)[1])
autoplot(train) + 
  autolayer(mean_model_d_levels) + 
  autolayer(test, alpha = 0.5) + 
  scale_color_manual(values = c('blue','red'),
                     label = c('actual','prediction'))

ets_model_d <- ets(train_diff)
ets_forecast_d <- forecast(ets_model_d, h = length(test), PI = FALSE)
ets_model_d_levels <- ts(cumsum(c(tail(train,1), ets_forecast_d$mean)),
                         frequency = frequency, start = time(test)[1])
autoplot(train, na.action = 'na.interp') + 
  autolayer(ets_model_d_levels) + 
  autolayer(test, alpha = 0.5) + 
  scale_color_manual(values = c('blue','red'),
                     label = c('actual','prediction'))
```

Check for outliers

```{r, echo = show_code}
ggplot(fortify(train), aes(x,y)) +
  geom_point(
    color = if_else(time(train) %in% tsoutliers(train)$index,
                    'red', 'black'))

ggplot(fortify(train_diff), aes(x,y)) +
  geom_point(
    color = if_else(time(train_diff) %in% tsoutliers(train_diff)$index,
                    'red', 'black'))

if (outlier_removal == TRUE) {
  train <- tsclean(train, replace.missing = FALSE)
  train_diff <- tsclean(train_diff, replace.missing = FALSE)
  cat('Outliers removed.')
}
```

Check for missing values

```{r, echo = show_code}
length(train[is.na(train)])

if (interpolate == TRUE) {
  if (missing_method == 'linear') {
    train <- zoo::na.approx(train)
    train_diff <- zoo::na.approx(train_diff)
  } else if (missing_method == 'spline') {
    train <- zoo::na.spline(train)
    train_diff <- zoo::na.spline(train_diff)
  }
  cat('Missing values replaced using',missing_method,'method')
}
```

Check for skewness

```{r, echo = show_code}
ggplot(fortify(train), aes(y)) +
  geom_histogram(bins = 50)

if (transform == TRUE) {
  if (trans_method == 'BoxCox') {
    train <- BoxCox(train, 
                    lambda = BoxCox.lambda(transform, method = 'loglik'))
  } else if (method == 'log') {
    train <- log(train)
  }
  cat('Series transformed using',trans_method)
}

ggplot(fortify(train_diff), aes(y)) +
  geom_histogram(bins = 50)

if (transform_diff == TRUE) {
  if (trans_method_diff == 'BoxCox') {
    train_diff <- BoxCox(train_diff, 
                         lambda = BoxCox.lambda(transform, method = 'loglik'))
  } else if (trans_method_diff == 'log') {
    train_diff <- log(train_diff)
  }
  cat('Differenced series transformed using',trans_method_diff)
}
```

Check for stationarity

```{r, echo = show_code}
summary(ur.kpss(train))

cat('\n################################',
    '\nSuggested number of differences:',
    ndiffs(train))
```

Check for seasonality

```{r, echo = show_code}
for (period in seasonal_periods) {
  seasonal_ts <- ts(train, frequency = period)
  cat('\nSuggested number of seasonal diffs for freq =',period,'-',
      nsdiffs(seasonal_ts))
  print(stl(seasonal_ts,
            s.window = period,
            t.window = period,
            robust = TRUE) %>%
        autoplot())
}
```

Check ACF/PACF

```{r, echo = show_code}
ggAcf(train)
ggPacf(train)

ggAcf(train_diff)
ggPacf(train_diff)
```

Manual ARIMA model fit

```{r, echo = show_code}
arima_model <- Arima(train, order = c(p,d,q), include.drift = drift)
summary(arima_model)
checkresiduals(arima_model)
arima_forecast <- forecast(arima_model, h = length(test))
autoplot(arima_forecast) +
  autolayer(test, alpha = 0.5)
```

Manual seasonal ARIMA fit

```{r, echo = show_code}
arima_model_s <- Arima(train, 
                       order = c(p,d,q), 
                       seasonal = c(p_s,d_s,q_s), 
                       include.drift = drift_s)
summary(arima_model_s)
checkresiduals(arima_model_s)
arima_forecast_s <- forecast(arima_model_s, h = length(test))
autoplot(arima_forecast_s) +
  autolayer(test, alpha = 0.5)
```

Auto ARIMA

```{r, echo = show_code}
autoarima_model <- auto.arima(train, 
                              approximation = approximation, 
                              stepwise = stepwise)
summary(autoarima_model)
checkresiduals(autoarima_model)
autoarima_forecast <- forecast(autoarima_model, h = length(test))
autoplot(autoarima_forecast) +
  autolayer(test, alpha = 0.5)
```

Evaluation

```{r, echo = show_code}
naive_mape <- mean(abs(naive_model$mean - test) / test, na.rm = TRUE)
naive_mape_d <- mean(abs(naive_model_d_levels - test) / test, na.rm = TRUE)
mean_mape <- mean(abs(mean_model$mean - test) / test, na.rm = TRUE)
mean_mape_d <- mean(abs(mean_model_d_levels - test) / test, na.rm = TRUE)
ets_mape <- mean(abs(ets_forecast$mean - test) / test, na.rm = TRUE)
ets_mape_d <- mean(abs(ets_model_d_levels - test) / test, na.rm = TRUE)

arima_mape <- mean(
  abs(arima_forecast$mean - test) / test, na.rm = TRUE)

arima_mape_s <- mean(
  abs(arima_forecast_s$mean - test) / test, na.rm = TRUE)

autoarima_mape <- mean(
  abs(autoarima_forecast$mean - test) / test, na.rm = TRUE)

results <- data.frame(
  model = c('naive','naive_d',
            'mean','mean_d',
            'ets','ets_d',
            'arima','seasonal arima', 'autoarima'),
  mape = c(naive_mape, naive_mape_d,
           mean_mape, mean_mape_d,
           ets_mape, ets_mape_d,
           arima_mape, arima_mape_s, autoarima_mape)
)

results %>%
  arrange(mape)
```

###
MANUAL - Generate Forecasts with chosen approach, using FULL SERIES
###

```{r}
# NAIVE
if (final_method == 'naive') {
  final_model <- naive(series, h = 140)
  final_forecast <- final_model$mean
  autoplot(naive_model)
}

# AVERAGE
if (final_method == 'average') {
  final_model <- meanf(series, h = 140)
  final_forecast <- final_model$mean
  autoplot(final_model)
}

# ETS
if (final_method == 'ets') {
  final_model <- ets(series, na.action = 'na.interp')
  final_forecast <- forecast(final_model, h = 140)
  autoplot(final_forecast)
}

# ARIMA
if (final_method == 'arima') {
  final_model <- Arima(series,
                       order = c(final_p,final_d,final_q),
                       seasonal = c(final_p_s,final_d_s,final_q_s),
                       include.drift = FALSE)
  final_forecast <- forecast(final_model, h = 140)
  autoplot(final_forecast)
}
```

Write final forecast to csv

```{r}
if (write_to_csv == TRUE) {
  filename <- paste0('data/finalForecast_',category, variable, '.csv')
  write.csv(final_forecast, filename, row.names = FALSE)
}
```

