---
title: 'Data 621 HW5: Wine Sales'
author: "(Group 4) Mohamed Hassan-El Serafi, Chun Shing Leung, Keith Colella, Yina Qiao, Eddie Xu"
date: "`r Sys.Date()`"
output: html_document
---

```{r, message=FALSE}
# Load required libraries
library(tidyverse) 
library(ggplot2) 
library(corrplot)
library(psych)
library(naniar)
library(cowplot)
library(mice)
library(pscl)
library(MASS)
library(randomForest)
library(knitr)
library(kableExtra)
```

# Data Exploration

```{r}
# Load the data
wine_data <- read.csv("data/wine-training-data.csv")

# Glimpse
glimpse(wine_data)
```

```{r}
eval_data <- read.csv("data/wine-evaluation-data.csv") %>%
  dplyr::select(-IN, -TARGET)
glimpse(eval_data)
```

### Summary

```{r}
# Drop index
wine_data <- wine_data %>%
  dplyr::select(-INDEX)

# Summary statistics for the dataset
summary_stats <- describe(wine_data)
print(summary_stats)
```

### NAs

```{r}
# Check for missing values
missing_values <- colSums(is.na(wine_data))
print(missing_values)
```

```{r}
# Missing value patterns
naniar_plot <- gg_miss_var(wine_data, show_pct = TRUE) +
  labs(title = "Missing Data Patterns", x = "Variables", y = "Number of Missing Values") +
  theme_minimal()

print(naniar_plot)

# Create a heatmap of missing values
gg_miss_upset(wine_data)
```

### Correlation across features

```{r}
# Correlation matrix for numerical variables
numerical_vars <- wine_data %>% select_if(is.numeric)
correlation_matrix_1 <- cor(numerical_vars, use = "pairwise.complete.obs")

# With log target
numerical_vars$TARGET <- log(numerical_vars$TARGET+1)
correlation_matrix_2 <- cor(numerical_vars, use = "pairwise.complete.obs")

# Plot the heatmap
corrplot(correlation_matrix_1, method = "color", type = "lower", title = "Untransformed Target", mar = c(1, 1, 2, 1))
corrplot(correlation_matrix_2, method = "color", type = "lower", title = "Log-transformed Target", mar = c(1, 1, 2, 1))
```

### Distributions

```{r}
# Histograms for numerical vars
histograms <- lapply(names(dplyr::select(numerical_vars, -TARGET)), function(var) {
  ggplot(wine_data, aes_string(x = var)) +
    geom_histogram()
})

# Arrange histograms in a grid
histogram_grid <- plot_grid(plotlist = histograms, ncol = 3)

# Save or display the plot grid
print(histogram_grid)
```

```{r}
# Histogram for target variable
ggplot(wine_data, aes(x = TARGET)) +
  geom_bar() +
  labs(title = "Distribution of TARGET (Cases Purchased)", x = "TARGET", y = "Frequency")
```

```{r}
variables <- names(wine_data)[!names(wine_data) %in% c("TARGET")]

plots <- lapply(variables, function(var) {
  ggplot(wine_data, aes_string(x = var, y = "TARGET")) +
    geom_point() 
})

# Arrange the plots in a grid (5 rows x 3 columns)
grid_plot <- plot_grid(plotlist = plots, nrow = 5, ncol = 3, rel_heights = c(2,2,2,2,2))

print(grid_plot)
```

```{r}
variables <- names(wine_data)[!names(wine_data) %in% c("TARGET")]

plots <- lapply(variables, function(var) {
  ggplot(wine_data, aes_string(x = var, y = "TARGET")) +
    geom_point() 
})

plots
```

# Evaluation Data

```{r}
# Remove all rows with missing values from the original data
complete_data <- na.omit(wine_data)

# Determine the number of rows you want in the validation set
set.seed(123) # for reproducibility
validation_size <- floor(0.2 * nrow(complete_data)) # 20% as an example

# Randomly select rows for the validation set
validation_indices <- sample(seq_len(nrow(complete_data)), size = validation_size)

# Create the validation subset with no NAs
validation_data <- complete_data[validation_indices, ]

# (Optional) Create a training subset from the remaining rows
wine_data <- wine_data[-validation_indices, ]

paste('Rows in Training Data', nrow(wine_data))
paste('Rows in Validation Data', nrow(validation_data))
```

# Modeling

```{r}
# Initialize an empty results data frame
results_df <- data.frame(
  ModelName = character(),
  RowsInTrain = numeric(),
  AIC = numeric(),
  MAE_Train = numeric(),
  MAE_Val = numeric(),
  Deviance = numeric(),
  Dispersion = numeric(),
  stringsAsFactors = FALSE
)

# Function for updating the df
update_results <- function(results_df, model, training_data, validation_data, name) {
  # Generate Predictions
  predicted_train <- predict(model, type = "response")
  actual_train <- training_data$TARGET
  
  predicted_val <- predict(model, newdata=validation_data, type = "response")
  actual_val <- validation_data$TARGET
  
  # Assign variables
  name <- name
  mae_train <- mean(abs(actual_train - predicted_train), na.rm = TRUE)
  mae_val <- mean(abs(actual_val - predicted_val), na.rm = TRUE)
  deviance <- model$deviance
  dispersion <- sum(residuals(model, type = "pearson")^2) / df.residual(model)
  
  # Error handling
  if (is.null(deviance)) deviance <- NA
  if (is.null(dispersion) || length(dispersion) == 0) dispersion <- NA
  
  aic <- tryCatch({
    AIC(model)
  }, error = function(e) {
    NA
  })
  
  # Add to df
  results_df <- rbind(results_df, data.frame(
    ModelName = name,
    RowsInTrain = nrow(training_data),
    AIC = aic,
    MAE_Train = mae_train,
    MAE_Val = mae_val,
    Deviance = deviance,
    Dispersion = dispersion,
    stringsAsFactors = FALSE
  ))
  
  return(results_df)
}
```

### Baseline - with NAs removed

```{r}
nrow(wine_data)
nrow(na.omit(wine_data))

wine_data_naomit <- na.omit(wine_data)
```

```{r}
# Fit a Poisson GLM with NAs omitted, as baseline
poisson_model_naomit <- glm(TARGET ~ ., family = poisson(link = "log"), data = wine_data_naomit)
summary(poisson_model_naomit)
```

```{r}
plot(poisson_model_naomit)
```

```{r}
results_df <- update_results(
  results_df, poisson_model_naomit, wine_data_naomit, validation_data, 'baseline poisson - NAs removed'
)
results_df
```

### Convert to factors

```{r}
# Replace NAs with 0 in STARS
wine_data[is.na(wine_data$STARS), 'STARS'] <- "Missing"

# Convert to factor
wine_data$STARS <- factor(wine_data$STARS)
wine_data$LabelAppeal <- factor(wine_data$LabelAppeal)

# Same for validation
validation_data[is.na(validation_data$STARS), 'STARS'] <- "Missing"
validation_data$STARS <- factor(validation_data$STARS, levels = levels(wine_data$STARS))
validation_data$LabelAppeal <- factor(validation_data$LabelAppeal, levels = levels(wine_data$LabelAppeal))

# Check number of rows with NAs
nrow(na.omit(wine_data))
```

```{r}
# Fit a Poisson GLM
poisson_model_factor <- glm(TARGET ~ . - 1, family = poisson(link = "log"), data = wine_data)
summary(poisson_model_factor)
```

```{r}
plot(poisson_model_factor)
```

```{r}
results_df <- update_results(
  results_df, poisson_model_factor, na.omit(wine_data), validation_data, 'poisson - discrete vars as factors'
)
results_df
```

### Impute Numeric vars

```{r}
wine_data_mean_imp <- wine_data

# Identify numeric columns
num_cols <- sapply(wine_data_mean_imp, is.numeric)

# For each numeric column with NAs, fill them with the column mean
for (col_name in names(wine_data_mean_imp)[num_cols]) {
  missing_indices <- is.na(wine_data_mean_imp[[col_name]])
  if (any(missing_indices)) {
    col_mean <- mean(wine_data_mean_imp[[col_name]], na.rm = TRUE)
    wine_data_mean_imp[[col_name]][missing_indices] <- col_mean
  }
}

# Fit a Poisson GLM
poisson_model_mean_impute <- glm(TARGET ~ . - 1, family = poisson(link = "log"), data = wine_data_mean_imp)
summary(poisson_model_mean_impute)
```

```{r}
results_df <- update_results(
  results_df, poisson_model_mean_impute, wine_data_mean_imp, validation_data, 'poisson - mean imputed'
)
results_df
```

```{r}
mice_imp_model <- mice(wine_data, m = 5, method = "pmm", seed = 123)
wine_data_mice_imp <- complete(mice_imp_model, action = 1)

poisson_model_mice_impute <- glm(TARGET ~ . - 1, family = poisson(link = "log"), data = wine_data_mice_imp)
summary(poisson_model_mice_impute)
```

```{r}
results_df <- update_results(
  results_df, poisson_model_mice_impute, wine_data_mice_imp, validation_data, 'poisson - mice imputed'
)
results_df
```

```{r}
wine_data <- wine_data_mice_imp
```

### Negative Binomial

```{r}
# Fit a Negative Binomial regression
nb_model <- glm.nb(TARGET ~ .  - 1, data = wine_data)
summary(nb_model)
```

```{r}
plot(nb_model)
```

```{r}
results_df <- update_results(
  results_df, nb_model, wine_data, validation_data, 'neg. binom. - mice imputed'
)
results_df
```

### Zero-Inflation

```{r}
zip_model <- zeroinfl(TARGET ~ . | 1, dist = "poisson", data = wine_data)
summary(zip_model)
```

```{r}
results_df <- update_results(results_df, zip_model, wine_data, validation_data, 'zip model - mice imputed')
results_df
```

### Additional Features

```{r}
# Identify numeric columns
num_cols <- sapply(wine_data, is.numeric)
vars <- names(wine_data)[num_cols]
vars <- setdiff(vars, "TARGET")

# Create a formula with all numeric variables, their pairwise interactions (^2),
# and squared polynomial terms (I(var^2))
f <- as.formula(
  paste("~ (", paste(vars, collapse=" + "), ")^2 +",
        paste(sapply(vars, function(v) paste0("I(", v, "^2)")), collapse=" + "))
)

# Generate the model matrix from the formula (remove the intercept column)
mm <- model.matrix(f, data = wine_data)[, -1]
mm_new_features <- mm[, !colnames(mm) %in% vars]

# Add the generated terms to the original data frame
wine_data_genFeatures <- data.frame(cbind(wine_data['TARGET'], mm_new_features))
names(wine_data_genFeatures) <- make.names(names(wine_data_genFeatures))

# Same for validation data
validation_mm <- model.matrix(f, data = validation_data)[, -1]
validation_new_features <- validation_mm[, !colnames(validation_mm) %in% vars]
validation_data_genFeatures <- data.frame(cbind(validation_data['TARGET'], validation_new_features))
names(validation_new_features) <- make.names(names(validation_new_features))
```

```{r}
# Stepwise elimination based on p-values
p_threshold <- 0.05
feature_selection_data <- wine_data_genFeatures

for (i in 1:(ncol(feature_selection_data)-1)) {
  # Fit model (no intercept)
  feature_selection_model <- glm(TARGET ~ . - 1, family = poisson, data = feature_selection_data)
  
  # Get p-values for all predictors (no intercept to remove)
  p_vals <- summary(feature_selection_model)$coefficients[, 4]
  
  # Identify the predictor with the highest p-value
  max_p <- max(p_vals, na.rm = TRUE)  # No [-1] since no intercept
  if (max_p < p_threshold) {
    message("STOPPING POINT!")
    break
  }
  
  # Find the predictor with the highest p-value
  worst_predictor <- names(p_vals)[which.max(p_vals)]
  
  # Remove the worst predictor by exact name, but only if not factor var
  worst_predictor <- make.names(worst_predictor)
  feature_selection_data <- dplyr::select(feature_selection_data, -all_of(worst_predictor))
}

summary(feature_selection_model)
```

### Add back factor variables

```{r}
wine_data_genFeatures <- cbind(feature_selection_data, dplyr::select(wine_data, STARS, LabelAppeal))
poisson_model_genFeatures <- glm(TARGET ~ . - 1, family = poisson, data = wine_data_genFeatures)
summary(poisson_model_genFeatures)
```

```{r}
plot(poisson_model_genFeatures)
```

```{r}
validation_data_genFeatures <- cbind(validation_data_genFeatures, dplyr::select(validation_data, STARS, LabelAppeal))

results_df <- update_results(
  results_df, poisson_model_genFeatures, wine_data_genFeatures, validation_data_genFeatures, 'poisson - generated features (with factors)'
)
results_df
```

### Negative Binomial

```{r}
# Fit a Negative Binomial regression
nb_model_genFeatures <- glm.nb(TARGET ~ .  - 1, data = wine_data_genFeatures)
summary(nb_model)
```

```{r}
plot(nb_model_genFeatures)
```

```{r}
results_df <- update_results(
  results_df, nb_model_genFeatures, wine_data_genFeatures, validation_data_genFeatures, 'neg. binom. - generated features (with factors)'
)
results_df
```

### Zero-Inflation

```{r}
zip_model_genFeatures <- zeroinfl(TARGET ~ . | 1, dist = "poisson", data = wine_data_genFeatures)
summary(zip_model_genFeatures)
```

```{r}
results_df <- update_results(results_df, zip_model_genFeatures, wine_data_genFeatures, validation_data_genFeatures, 'zip model - generated features (with factors)')
results_df
```

### Linear Regression

```{r}
lr_model <- lm(TARGET ~ . - 1, data = wine_data)
summary(lr_model)
```

```{r}
plot(lr_model)
```

```{r}
results_df <- update_results(results_df, lr_model, wine_data, validation_data, 'linear regression - mice imputed')
results_df
```

```{r}
wine_data_log <- wine_data %>%
  mutate(TARGET = log(TARGET+1))
lr_model_log <- lm(TARGET ~ . - 1, data = wine_data_log)
summary(lr_model_log)
```

```{r}
plot(lr_model_log)
```

```{r}
validation_data_log <- validation_data %>%
  mutate(TARGET = log(TARGET+1))

predicted_train <- predict(lr_model_log)
predicted_lin_space_train <- exp(predicted_train) - 1
mae_train <- mean(abs(wine_data$TARGET - predicted_lin_space_train))

predicted_val <- predict(lr_model_log, newdata=validation_data_log)
predicted_lin_space_val <- exp(predicted_val) - 1
mae_val <- mean(abs(validation_data$TARGET - predicted_lin_space_val))

results_df <- rbind(results_df, data.frame(
  ModelName = 'linear regression - log target',
  RowsInTrain = nrow(wine_data_log),
  AIC = NA,
  MAE_Train = mae_train,
  MAE_Val = mae_val,
  Deviance = NA,
  Dispersion = NA,
  stringsAsFactors = FALSE
))

results_df
```

### Random Forest

```{r}
rf_model <- randomForest(TARGET ~ ., data = wine_data, ntree = 500)
summary(rf_model)
```


```{r}
results_df <- update_results(results_df, rf_model, wine_data, validation_data, 'rf model - mice imputed')
results_df
```

### RF with Log TARGET

```{r}
rf_model_log <- randomForest(TARGET ~ ., data = wine_data_log, ntree = 500)
summary(rf_model_log)
```

```{r}
validation_data_log <- validation_data %>%
  mutate(TARGET = log(TARGET+1))

predicted_train <- predict(rf_model_log)
predicted_lin_space_train <- exp(predicted_train) - 1
mae_train <- mean(abs(wine_data$TARGET - predicted_lin_space_train))

predicted_val <- predict(rf_model_log, newdata=validation_data_log)
predicted_lin_space_val <- exp(predicted_val) - 1
mae_val <- mean(abs(validation_data$TARGET - predicted_lin_space_val))

results_df <- rbind(results_df, data.frame(
  ModelName = 'rf model - log target',
  RowsInTrain = nrow(wine_data_log),
  AIC = NA,
  MAE_Train = mae_train,
  MAE_Val = mae_val,
  Deviance = NA,
  Dispersion = NA,
  stringsAsFactors = FALSE
))

results_df
```

# Summary Table

```{r}
results_df %>%
  kbl(
    caption = "Summary of Model Performance",
    col.names = c("Model Name", "No. Training Observations", "AIC", "MAE (Train)", "MAE (Validation)", "Deviance", "Dispersion"),
    digits = 2
  ) %>%
  kable_classic(position = 'center')
```

# Final Predictions

```{r}
# Load in final data
final_eval_data <- read.csv('data/wine-evaluation-data.csv')

# Replace NAs with 0 in STARS
final_eval_data[is.na(final_eval_data$STARS), 'STARS'] <- "Missing"

# Convert to factor
final_eval_data$STARS <- factor(final_eval_data$STARS)
final_eval_data$LabelAppeal <- factor(final_eval_data$LabelAppeal)

final_eval_data_imputed <- complete(mice(final_eval_data, m = 5, method = "pmm", seed = 123), 1) # complete(mice.mids(predict.mids(mice_imp_model, newdata = final_eval_data)), 1)

final_predictions_raw <- predict(zip_model, newdata=final_eval_data_imputed)
final_predictions_round <- floor(final_predictions_raw)

predictions_df <- data.frame(cbind(final_predictions_raw, final_predictions_round))
write.csv(predictions_df, file='hw5_predictions.csv', row.names=FALSE)
```

# Visualize predictions aross subsets.

```{r}
train_predictions <- predict(zip_model)
val_predictions <- predict(zip_model, newdata=validation_data)

hist(floor(train_predictions))
hist(floor(val_predictions))
hist(floor(final_predictions_raw))
```

