---
title: "Visualizations"
author: "John Cruz"
date: "3-16-2022"
output: html_document
---

```{r setup, message=FALSE}
library(tidyverse)
library(ggplot2)
library(ggwordcloud)
```

## Read in data

Read the data from a CSV file which contains job listings and skills

```{r message=FALSE}
import_df <- read_csv('output/job_listings_skills_long.csv')
```

## Job Listings Based on Data Source

```{r website-count}
distinct_job_website <- 
  import_df |> 
  group_by(website) |> 
  summarise(count = n_distinct(job_id)) |> 
  arrange(desc(count))

ggplot(distinct_job_website, aes(x = count, y = reorder(website, count))) +
  geom_bar(stat = 'identity', fill = '#4E79A7') +
  labs(title = "Job Listings per Website", x = '', y = '') +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text = element_text(face = 'bold'))

knitr::kable(distinct_job_website)
```

## Most Requested Degree

Based on the job postings that had college degree information, having a Master's was first at 498, followed by a Bachelor's at 457. 

```{r college-degree}
high_degree <- 
  import_df |> 
  na.omit(highest_ed) %>%
  group_by(highest_ed) %>%
  summarize(count = n_distinct(job_id))

high_degree %>%
  ggplot(aes(x = highest_ed, y = count)) +
  geom_bar(stat = 'identity', fill = '#4E79A7') + 
  labs(title = "Most Degrees Requested", x = '', y = '') +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        axis.text = element_text(face = 'bold')) +
  geom_text(aes(label = count), 
            position = position_stack(vjust = 0.8), 
            size = 4,
            fontface = 'bold')

knitr::kable(high_degree)
```

## Maximum Salary Distribution

The histogram shows the distribution of maximum salary values for all jobs within the dataset, with missing values removed. We can see that there is a bi-modal distribution where it peaks first at \$125,000 with 1,365 jobs and again at \$175,000 with 986 jobs. 

```{r salary}
salaries <- 
  import_df |> 
  filter(!is.na(salary_max)) |> 
  distinct(job_id, .keep_all = TRUE)
  

salaries |> 
  ggplot(aes(x = salary_max)) +
  geom_histogram(binwidth = 25000, fill = '#4E79A7') +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Maximum Salary Distribution", x = '', y = '') +
  theme_bw() +
  stat_bin(binwidth=25000, geom='text', color='white', size=3,
           aes(label=after_stat(count)), position=position_stack(vjust=0.8))
```

## Maximum Salary Distribution vs. Education

```{r rename}
salaries <- 
  import_df |> 
  filter(!is.na(salary_max)) |> 
  distinct(job_id, .keep_all = TRUE)
  

salaries |> 
  ggplot(aes(x = salary_max)) +
  geom_histogram(binwidth = 25000, fill = '#4E79A7') +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Maximum Salary Distribution", x = '', y = '') +
  theme_bw() +
  stat_bin(binwidth=25000, geom='text', color='white', size=3,
           aes(label=after_stat(count)), position=position_stack(vjust=0.8))
```

## Title

```{r rename2}

```

## Title

```{r rename3}

```

## Title

```{r rename4}
explore <- read_csv('output/job_listings.csv')
explore |> 
  group_by(website) |> 
  summarise(count = n())
```












