---
title: "CUNY SPS MSDS Bridge Program - R - Week 2 Homework"
author: "Keith Colella"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

I'll use the carData package, focusing on the Arrests dataset, which details arrests in Toronto for simple possession of small quantities of marijuana.

```{r}
# install.packages("carData")
library(carData)
 head(Arrests,5)
```


# Question #1 - Summary of Data

```{r}
summary(Arrests)
cat('Mean Age:', mean(Arrests$`age`),'\n',
    'Median Age:', median(Arrests$`age`), '\n')
cat('Mean Checks:', mean(Arrests$`checks`),'\n',
    'Median Checks:', median(Arrests$`checks`), '\n')
```

# Question #2 - Subset of dataset

```{r}
male_post2000 <- subset(Arrests,
                        sex = 'Male',
                        year >= 2000,
                        select = c(released,colour,year,age,sex,employed,checks)
                        )
head(male_post2000,5)
```

# Question #3 - New column names

```{r}
library(plyr)
male_post2000 <- rename(male_post2000, c('released' = 'Release (Y/N)', 
                                         'colour' = 'Race', 
                                         'checks' = 'No. of Checks'
                                         )
                        )

head(male_post2000,5)
```

# Question #4 - Summary of new dataframe / Comparison to original dataset

```{r}
summary(male_post2000)
```
First, we compare the mean / median Age between the original and subset datasets.
```{r}
cat('Original Mean:', mean(Arrests$age),'\n',
    'Subset Mean:', mean(male_post2000$age), '\n')
cat('Original Median:',median(Arrests$age),'\n',
    'Subset Median:',median(male_post2000$age),'\n')
```

Next, we compare the mean / median No. of checks between each dataset.
```{r}
cat('Original Mean:', mean(Arrests$checks),'\n',
    'Subset Mean:', mean(male_post2000$`No. of Checks`), '\n')
cat('Original Median:',median(Arrests$checks),'\n',
    'Subset Median:',median(male_post2000$`No. of Checks`),'\n')
```

The subsets have only minimally different means, and the same medians. This indicates that the subset largely resembles the original data set.

# Question #5 - Adjusting / Renaming Values for Three Columns

For a column with factors, we can change the levels.
```{r}
levels(male_post2000$employed)
levels(male_post2000$employed) <- c('Y','N')
```

For a column with numerical data, we can perform an operation.
```{r}
male_post2000$age <- male_post2000$age + 5
```

As this dataset has no character datatypes (only factors), I first changed the datatype to character. Then we can rename the values.
```{r}
male_post2000$`Release (Y/N)` = as.character(male_post2000$`Release (Y/N)`)
male_post2000['Release (Y/N)'][male_post2000['Release (Y/N)'] == 'Yes'] <- 'Y'
male_post2000['Release (Y/N)'][male_post2000['Release (Y/N)'] == 'No'] <- 'N'
```

# Question #6 - View Changes

```{r}
head(male_post2000,10)
```

# BONUS

I downloaded the original Arrests dataset, then uploaded to my github repository, from which the code below then reads in the file.
```{r}
Arrests2 <- read.csv('https://raw.githubusercontent.com/kac624/notebooks/main/CUNY/Bridge/Arrests.csv')
head(Arrests2,5)
```

