---
title: "s02v02"
author: "Keith Colella"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup}
library(tidyverse)
library(readxl)
library(forecast)
library(urca)
```

Choice of variable

```{r}
category <- 'S02'
```


```{r}
variable <- 'Var03'
```

Data

```{r}
xl <- read_excel('data/DATA624_Project1_Data_Schema.xlsx',
                 sheet = category, skip = 2)

series <- xl[1:1622,variable]
```

Convert to ts object and plot. Assume freq = 1 for now.

```{r}
series <- ts(deframe(series), frequency = 1)
autoplot(series)
```

Split training and test sets

```{r}
train <- head(series, round(length(series) * 0.7))
test <- tail(series, round(length(series) * 0.3))

cat('Obs. in test set:',length(train),
    '\nObs. in test set:',length(test),
    '\nTotal observations:',length(series))
    
autoplot(train)
autoplot(test)
```

Split training and test sets with differenced series

```{r}
train_diff <- diff(train)
test_diff <- diff(test)

autoplot(train_diff)
autoplot(test_diff)
```

Baseline forecasts - Naive, Average and ETS

```{r}
naive_model <- naive(train, h = length(test), level = 0)
autoplot(naive_model) + 
  autolayer(test, alpha = 0.5)

mean_model <- meanf(train, h = length(test), level = 0)
autoplot(mean_model) + 
  autolayer(test, alpha = 0.5)

ets_model <- ets(train)
ets_forecast <- forecast(ets_model, h = length(test), PI = FALSE)
autoplot(ets_forecast) + 
  autolayer(test, alpha = 0.5)
```

Baseline forecasts on differenced series

```{r}
naive_model_d <- naive(train_diff, h = length(test), level = 0)
autoplot(naive_model_d) + 
  autolayer(test_diff, alpha = 0.5)

mean_model_d <- meanf(train_diff, h = length(test), level = 0)
autoplot(mean_model_d) + 
  autolayer(test_diff, alpha = 0.5)

ets_model_d <- ets(train_diff)
ets_forecast_d <- forecast(ets_model_d, h = length(test), PI = FALSE)
autoplot(ets_forecast_d) + 
  autolayer(test_diff, alpha = 0.5)
```

Check for skewness

```{r}
ggplot(data.frame(data = train)) +
  geom_histogram(aes(data), bins = 50)

ggplot(data.frame(data = train_diff)) +
  geom_histogram(aes(data), bins = 50)
```

Check for seasonality

```{r}
for (period in c(5,20,252)) {
  print(stl(ts(train, frequency = period),
            s.window = period,
            t.window = 250,
            robust = TRUE) %>%
        autoplot())
}
```

Check for outliers

```{r}
length(boxplot.stats(train, coef = 2.5)$out)
length(boxplot.stats(train_diff, coef = 2.5)$out)
```

Check for stationarity

```{r}
ur.kpss(train) %>% summary()
```

Check ACF/PACF

```{r}
ggAcf(train)
ggPacf(train)

ggAcf(train_diff)
ggPacf(train_diff)
```

Manual ARIMA model fit

```{r}
arima_model <- Arima(train, order = c(1,1,1), include.drift = TRUE)
summary(arima_model)
checkresiduals(arima_model)
arima_forecast <- forecast(arima_model, h = length(test))
autoplot(arima_forecast) +
  autolayer(test)
```

Auto ARIMA

```{r}
autoarima_model <- auto.arima(train, 
                              approximation = FALSE, 
                              stepwise = FALSE)
summary(autoarima_model)
checkresiduals(autoarima_model)
autoarima_forecast <- forecast(autoarima_model, h = length(test))
autoplot(autoarima_forecast) +
  autolayer(test)
```

Seasonal Modeling

```{r}
# TBD
```

Evaluation

```{r}
naive_mape <- mean(abs(naive_model$mean - test) / test, na.rm = TRUE)
naive_mape_d <- mean(abs(naive_model_d$mean - test) / test, na.rm = TRUE)
mean_mape <- mean(abs(mean_model$mean - test) / test, na.rm = TRUE)
mean_mape_d <- mean(abs(mean_model_d$mean - test) / test, na.rm = TRUE)
ets_mape <- mean(abs(ets_forecast$mean - test) / test, na.rm = TRUE)
ets_mape_d <- mean(abs(ets_forecast_d$mean - test) / test, na.rm = TRUE)

arima_mape <- mean(
  abs(arima_forecast$mean - test) / test, na.rm = TRUE)

autoarima_mape <- mean(
  abs(autoarima_forecast$mean - test) / test, na.rm = TRUE)

results <- data.frame(
  model = c('naive','naive_d',
            'mean','mean_d',
            'ets','ets_d',
            'arima','autoarima'),
  mape = c(naive_mape, naive_mape_d,
           mean_mape, mean_mape_d,
           ets_mape, ets_mape_d,
           arima_mape, autoarima_mape)
)

results %>%
  arrange(mape)
```

